\bf Eye Fitting Straight Lines in the Modern Era

### 2021-09-10 Eye-Fitting-Straight-Lines-in-the-Modern-Era.tex
rough draft,. still need to add intro though...

\begin{itemize}
\tightlist
\item
  Cleveland and McGill
\end{itemize}

@@

Graphical tests are useful for studying the perception of statistical
graphs. Studies might ask participants to identify differences in
graphs, read information off of a chart accurately, use data to make
correct real-world decisions, or predict the next few observations. All
of these types of tests require different levels of use and manipulation
of the information being presented in the chart. Early researchers
studied graphs from a psychological perspective. These studies generally
tested participants ability to detect a stimulus (or a difference
between two stimuli) ({\textcolor{blue}{CITATIONS}}). Here we focus on
the how graphical testing has developed in statistics.

@@

One useful tool for testing statistical graphics is the concept of a
lineup. \citet{buja2009statistical} introduced the lineup protocol in
which data plots are depicted and interpreted as statistics. Supported
by the grammar of graphics, a data plot can be characterized as a
statistic, defined as, ``a functional mapping of a variable or set of
variables'' \citep{vanderplas2020testing}. This allows the data plot to
be tested similar to other statistics, by comparing the actual data plot
to a set of plots with the absence of any data structure we can test the
likelihood of any perceived structure being significant. The
construction of data plots as statistics allow for easy experimentation,
granting researchers the ability to compare the effectiveness of and
understand the perception of different types of charts
({\textcolor{blue}{CITATIONS}}). The lineup protocol is one such example
of the development of tools designed for statistical graphical testing.
The advancement of graphing software provides the tools necessary to
develop new methods of testing graphics.

@@

\citet{mosteller1981eye}, sought to understand the properties of least
squares and other computed lines by establishing one systematic method
of fitting lines by eye. Participants were asked to fit lines by eye to
four scatter-plots using an 8.5 x 11 inch transparency with a straight
line etched completely across the middle. A latin square design with
packets of the set of points stapled together in four different
sequences was used to determine if there is an effect of order of
presentation. It was found that order of presentation had no effect and
that participants tended to fit the slope of the principal axis (error
minimized orthogonally, both horizontal and vertical, to the regression
line) over the slope of the least squares regression line (error
minimized vertically to the regression line). These results support
previous research on ``ensemble perception'' indicating the visual
system can compute averages of various features in parallel across the
items in a set ({\textcolor{blue}{CITATIONS}}).

@@

In this paper, we establish `You Draw It', adapted from the New York
Times feature, as tool for graphical testing. The `You Draw It' method
is validated by replicating the study conducted by
\citet{mosteller1981eye}. The results of our study provide support for
``ensemble perception''.

@@

Data Driven Documents (D3), a JavaScript-based graphing framework that
facilitates user interaction, is used to create the `You Draw It' task
plots. Integrating this into RShiny using the r2d3 package, participants
are asked to draw a line using their computer mouse through trend shown
on their screen. Details of the development of the `You Draw It' task
plots will be addressed in future work. In the study, participants are
shown an interactive scatter-plot \cref{fig:ydi-stimuli} along with the
prompt, ``Use your mouse to fill in the trend in the yellow box
region.'' The yellow-box region moves along as the participant draws
their trend-line until the yellow region disappears, indicating the
participant has filled in the entire domain.

@@

\caption{'You Draw It' Stimuli}\label{fig:ydi-stimuli}
\end{figure}

@@

All data processing was conducted in R statistical software. A total of
\(N = 30\) points \((x_i, y_i), i = 1,...N\) were generated for
\(x_i \in [x_{min}, x_{max}]\) where \(x\) and \(y\) have a linear
relationship. Data were simulated based on linear model with additive
errors: \begin{align}
y_i & = \beta_0 + \beta_1 x_i + e_i \\
\text{with } e_i & \sim N(0, \sigma^2). \nonumber
\end{align} The parameters \(\beta_0\) and \(\beta_1\) are selected to
replicate \citet{mosteller1981eye} with \(e_i\) generated by rejection
sampling in order to guarantee the points shown align with that of the
fitted line.

Simulated model equation parameters were selected to reflect the four
data sets (F, N, S, and V) used in \citet{mosteller1981eye}
(\cref{tab:eyefitting-parameters}). Parameter choices F, N, and S
simulated data across a domain of 0 to 20. Parameter choice F produces a
trend with a positive slope and a large variance while N has a negative
slope and a large variance. In comparison, S shows a trend with a
positive slope with a small variance and V yields a steep positive slope
with a small variance over the domain of 4 to 16.
\cref{fig:eyefitting-simplot} illustrates an example of simulated data
for all four parameter choices intended to reflect the trends in
\citet{mosteller1981eye}. Aesthetic design choices were made consistent
across each of the interactive you draw it plots. The y-axis range
extended 10\% beyond (above and below) the range of the simulated data
points to allow for users to draw outside the simulated data set range.

@@

\caption{\label{tab:eyefitting-parameters}Eye Fitting Straight Lines in the Modern Era simulation model parameters}
\centering
\begin{tabular}[t]{cccc}
\toprule
Parameter Choice & $y_{\bar{x}}$ & $\beta_1$ & $\sigma$\\
\midrule
S & 3.88 & 0.66 & 1.30\\
F & 3.90 & 0.66 & 1.98\\
V & 3.89 & 1.98 & 1.50\\
N & 4.11 & -0.70 & 2.50\\
\bottomrule
\end{tabular}
\end{table}

@@

\caption{Eye Fitting Straight Lines in the Modern Era Simulated Data Example}\label{fig:eyefitting-simplot}
\end{figure}

@@

This experiment was conducted as part of a larger study; for simplicity,
we focus on the study design and methods related to the current study.
Each scatter-plot was the graphical representation of a data set that
was generated randomly, independently for each participant at the start
of the experiment. Participants in the study are shown two `You Draw It'
practice plots in order to train participants in the skills associated
with executing the task followed by four task plots associated with the
current study. The order of the task plots was randomly assigned for
each individual in a completely randomized design.

@@

We compare the participant drawn line to two regression lines determined
by ordinary least squares regression and regression based on the
principal axis (i.e.~Deming Regression). \cref{fig:ols-vs-pca-example}
illustrates the difference between an OLS regression equation which
minimizes the vertical distance of points from the line and a regression
equation based on the principal axis which minimizes the Euclidean
distance of points (orthogonal) from the line.

Due to the randomness in the data generation process, the actual slope
of the linear regression line fit through the simulated points could
differ from the predetermined slope. Therefore, we fit an ordinary least
squares (OLS) regression to each scatter-plot to obtain fitted values in
0.25 increments across the domain, \(\hat y_{k,OLS}\) for
\(k = 1, ..., 4 x_{max} +1\).

The regression equation based on the principal axis was determined by
using the \texttt{princomp} function in the stats package in base R to
obtain the rotation of the coordinate axes from the first principal
component (direction which captures the most variance). The estimated
slope, \(\hat\beta_{1,PCA}\), is determined by the ratio of the axis
rotation in y and axis rotation in x of the first principal component
with the y-intercept, \(\hat\beta_{0,PCA}\) calculated by the
point-slope equation of a line using the mean of of the simulated
points, \((\bar x_i, \bar y_i)\). Fitted values, \(\hat y_{k,PCA}\), are
then obtained every 0.25 increment across the domain from the PCA
regression equation,
\(\hat y_{k,PCA} = \hat\beta_{0,PCA} + \hat\beta_{1,PCA} x_k\).

@@

{\centering \includegraphics[width=1\linewidth,]{Eye-Fitting-Straight-Lines-in-the-Modern-Era_files/figure-latex/ols-vs-pca-example-1}

@@

\caption{OLS vs PCA Regression Lines}\label{fig:ols-vs-pca-example}
\end{figure}

\hypertarget{data-availability}{%
\subsection{Data Availability}\label{data-availability}}

For each participant, the final data set used for analysis contains
\(x_{ijk}, y_{ijk,drawn}, \hat y_{ijk,OLS}\), and \(\hat y_{ijk,PCA}\)
for parameter choice \(i = 1,2,3,4\), j = \(1,...N_{participant}\), and
\(x_{ijk}\) value \(k = 1, ...,4 x_{max} + 1\). Using both a linear
mixed model and a generalized additive mixed model, comparisons of
vertical residuals in relation to the OLS fitted values
(\(e_{ijk,OLS} = y_{ijk,drawn} - \hat y_{ijk,OLS}\)) and PCA fitted
values (\(e_{ijk,PCA} = y_{ijk,drawn} - \hat y_{ijk,PCA}\)) were made
across the domain. \cref{fig:eyefitting-example-plot} displays an
example of all three fitted trend lines for parameter choice F.

@@

{\centering \includegraphics[width=0.65\linewidth,]{Eye-Fitting-Straight-Lines-in-the-Modern-Era_files/figure-latex/eyefitting-example-plot-1}

@@

\caption{Eye Fitting Straight Lines in the Modern Era Example}\label{fig:eyefitting-example-plot}
\end{figure}

\hypertarget{linear-trend-constraint}{%
\subsection{Linear Trend Constraint}\label{linear-trend-constraint}}

Using the lmer function in the lme4 package \citep{lme4}, a linear mixed
model (LMM) is fit separately to the OLS and PCA residuals, constraining
the fit to a linear trend. Parameter choice, \(x\), and the interaction
between \(x\) and parameter choice were treated as fixed effects with a
random participant effect accounting for variation due to participant.
The LMM equation for each fit (OLS and PCA) residuals is given by:
\begin{equation}
y_{ijk,drawn} - \hat y_{ijk,fit} = e_{ijk,fit} = \left[\gamma_0 + \alpha_i\right] + \left[\gamma_{1} x_{ijk} + \gamma_{2i} x_{ijk}\right] + p_{j} + \epsilon_{ijk}
\end{equation} \noindent where

@@

@@

\hypertarget{smoothing-spline-trend}{%
\subsection{Smoothing Spline Trend}\label{smoothing-spline-trend}}

Eliminating the linear trend constraint, the bam function in the mgcv
package \citep{mgcv1, mgcv2, mgcv3, mgcv4, mgcv5} is used to fit a
generalized additive mixed model (GAMM) separately to the OLS and PCA
residuals to allow for estimation of smoothing splines. Parameter choice
was treated as a fixed effect with no estimated intercept and a separate
smoothing spline for \(x\) was estimated for each parameter choice. A
random participant effect accounting for variation due to participant
and a random spline for each participant accounted for variation in
spline for each participant. The GAMM equation for each fit (OLS and
PCA) residuals is given by: \begin{equation}
y_{ijk, drawn} - \hat y_{ijk, fit} = e_{ijk,fit} = \alpha_i + s_{i}(x_{ijk}) + p_{j} + s_{j}(x_{ijk})
\end{equation} \noindent where

@@

\cref{fig:eyefitting-lmer-residualplots} and
\cref{fig:eyefitting-gamm-residualplots} show the estimated trends of
residuals (vertical deviation of participant drawn points from both the
OLS and PCA fitted points) as modeled by a LMM and GAMM respectively.
Examining the plots, the estimated trends of PCA residuals (orange)
appear to align closer to the \(y=0\) horizontal (dashed) line than the
OLS residuals (blue). In particular, this trend is more prominent in
parameter choices with large variances (F and N). These results are
consistent to those found in \citet{mosteller1981eye} indicating
participants fit a trend line closer to the estimated regression line
with the slope of the first principal component than the estimated OLS
regression line.

@@

The intent of research was to establish you draw it as a tool for
testing graphics then use this tool to determine the cognitive
implications of displaying data on the log scale. Eye Fitting Straight
Lines in the Modern Era replicated the results found in
\citet{mosteller1981eye}. When shown points following a linear trend,
participants tended to fit the slope of the first principal component
over the slope of the least squares regression line. This trend was most
prominent when shown data simulated with larger variances. The
reproducibility of these results serve as evidence of the reliability of
the you draw it method.

@@

This study provided a basis for the use of `You Draw It' as a tool for
testing graphics. Further investigation is necessary to implement the
method in non-linear settings and with real data. This tool could also
be used to evaluate human ability to extrapolate data from trends. In
the future, an R package designed for easy implementation of `You Draw
It' task plots will help make this tool accessible to other researchers.

>>>>

\begin{itemize}
\tightlist
\item
  \citet{cleveland1984graphical}
\item
  \citet{cleveland1985graphical}
\end{itemize}

@@

Graphical tests are useful for studying the perception of statistical
graphs. Studies might ask participants to identify differences in
graphs, read information off of a chart accurately, use data to make
correct real-world decisions, or predict the next few observations. All
of these types of tests require different levels of use and manipulation
of the information being presented in the chart. Early researchers
studied graphs from a psychological perspective
\citep{spence1990visual, lewandowsky1989perception}. These studies
generally tested participants ability to detect a stimulus or a
difference between two stimuli. Here we focus on the how graphical
testing has developed in statistics.

@@

One useful tool for testing statistical graphics is the concept of a
lineup. \citet{buja2009statistical} introduced the lineup protocol in
which data plots are depicted and interpreted as statistics. Supported
by the grammar of graphics, a data plot can be characterized as a
statistic, defined as, ``a functional mapping of a variable or set of
variables'' \citep{vanderplas2020testing}. This allows the data plot to
be tested similar to other statistics, by comparing the actual data plot
to a set of plots with the absence of any data structure we can test the
likelihood of any perceived structure being significant. The
construction of data plots as statistics allow for easy experimentation,
granting researchers the ability to compare the effectiveness of and
understand the perception of different types of charts
\citep{vanderplas2017clusters, vanderplas2015spatial, hofmann2012graphical}.
The lineup protocol is one such example of the development of tools
designed for statistical graphical testing. The advancement of graphing
software provides the tools necessary to develop new methods of testing
graphics.

@@

\citet{mosteller1981eye}, sought to understand the properties of least
squares and other computed lines by establishing one systematic method
of fitting lines by eye. Participants were asked to fit lines by eye to
four scatter-plots using an 8.5 x 11 inch transparency with a straight
line etched completely across the middle. A latin square design with
packets of the set of points stapled together in four different
sequences was used to determine if there is an effect of order of
presentation. It was found that order of presentation had no effect and
that participants tended to fit the slope of the principal axis (error
minimized orthogonally, both horizontal and vertical, to the regression
line) over the slope of the least squares regression line (error
minimized vertically to the regression line). These results support
previous research on ``ensemble perception'' indicating the visual
system can compute averages of various features in parallel across the
items in a set
\citep{chong2003representation, chong2005statistical, van2011rapid}.

@@

In this paper, we establish `You Draw It', adapted from the New York
Times feature, as a tool for graphical testing. The `You Draw It' method
is validated by replicating the study conducted by
\citet{mosteller1981eye}. Based on previous research surrounding
``ensemble perception,'' we hypothesize that regression lines fit by
human perception resemble regression lines based on the principal axis
rather than an ordinary least squares regression line.

@@

Data Driven Documents (D3), a JavaScript-based graphing framework that
facilitates user interaction, is used to create the `You Draw It' task
plots. Integrating this into RShiny using the \texttt{r2d3} package,
participants are asked to draw a trend-line using their computer mouse
through a scatter-plot shown on their screen. In the study, participants
are shown an interactive scatter-plot (\cref{fig:ydi-stimuli}) along
with the prompt, ``Use your mouse to fill in the trend in the yellow box
region.'' The yellow box region moves along as the participant draws
their trend-line until the yellow region disappears, indicating the
participant has filled in the entire domain. Details of the development
of the `You Draw It' task plots will be addressed in future work.

@@

\caption{'You Draw It' task plot as shown to particpants during the study. The first frame (left) illustrates what particpants first see with the prompt “Use your mouse to fill in the trend in the yellow box region.” The second frame (middle), illustrates what the particpant sees while completing the task; the yellow region provides a visual cue for participants indicating where the participant still needs to complete a trend-line. The last frame (right) illustrates the participants finished trend-line before submission.}\label{fig:ydi-stimuli}
\end{figure}

@@

All data processing was conducted in R statistical software. A total of
\(N = 30\) points \((x_i, y_i), i = 1,...N\) were generated for
\(x_i \in [x_{min}, x_{max}]\) where \(x\) and \(y\) have a linear
relationship. Data were simulated based on a linear model with additive
errors: \begin{align}
y_i & = \beta_0 + \beta_1 x_i + e_i \\
\text{with } e_i & \sim N(0, \sigma^2). \nonumber
\end{align}

Model equation parameters, \(\beta_0\) and \(\beta_1\), were selected to
reflect the four data sets (F, N, S, and V) used in
\citet{mosteller1981eye} (\cref{tab:eyefitting-parameters}). Parameter
choices F, N, and S simulated data across a domain of 0 to 20. Parameter
choice F produces a trend with a positive slope and a large variance
while N has a negative slope and a large variance. In comparison, S
shows a trend with a positive slope with a small variance and V yields a
steep positive slope with a small variance over the domain of 4 to 16.
\cref{fig:eyefitting-simplot} illustrates an example of simulated data
for all four parameter choices intended to reflect the trends in
\citet{mosteller1981eye}. Aesthetic design choices were made consistent
across each of the interactive `You Draw It' task plots. The y-axis
range extended 10\% beyond (above and below) the range of the simulated
data points to allow for users to draw outside the simulated data set
range.

@@

\caption{\label{tab:eyefitting-parameters}Designated model equation parameters for simulated data.}
\centering
\begin{tabular}[t]{cccc}
\toprule
Parameter Choice & $y_{\bar{x}}$ & $\beta_1$ & $\sigma$\\
\midrule
S & 3.88 & 0.66 & 1.30\\
F & 3.90 & 0.66 & 1.98\\
V & 3.89 & 1.98 & 1.50\\
N & 4.11 & -0.70 & 2.50\\
\bottomrule
\end{tabular}
\end{table}

@@

\caption{Example of simulated data points displayed in a scatter-plot illustrating the trends associated with the four selected parameter choices.}\label{fig:eyefitting-simplot}
\end{figure}

@@

This experiment was conducted as part of a larger study; for simplicity,
we focus on the study design and methods related to the current study.
Each scatter-plot was the graphical representation of a data set that
was generated randomly, independently for each participant at the start
of the experiment. Participants in the study are shown two `You Draw It'
practice plots in order to train participants in the skills associated
with executing the task. During the practice session, participants are
provided with instruction prompts accompanied by a .gif and a practice
plot. Instructions guide participants to start at the edge of the yellow
box, to make sure the yellow region is moving along with their mouse as
they draw, and that they can draw over their already drawn line.
Practice plots are then followed by four `You Draw It' task plots
associated with the current study. The order of the task plots was
randomly assigned for each individual in a completely randomized design.

@@

We compare the participant drawn line to two regression lines determined
by ordinary least squares regression and regression based on the
principal axis (i.e.~Deming Regression). \cref{fig:ols-vs-pca-example}
illustrates the difference between an OLS regression line which
minimizes the vertical distance of points from the line and a regression
line based on the principal axis which minimizes the Euclidean distance
of points (orthogonal) from the line.

Due to the randomness in the data generation process, the actual slope
of the linear regression line fit through the simulated points could
differ from the predetermined slope. Therefore, we fit an ordinary least
squares (OLS) regression to each scatter-plot to obtain estimated
parameters \(\hat\beta_{0,OLS}\) and \(\hat\beta_{1,OLS}\). Fitted
values, \(\hat y_{k,OLS}\), are then obtained every 0.25 increments
across the domain from the OLS regression equation,
\(\hat y_{k,OLS} = \hat\beta_{0,OLS} + \hat\beta_{1,OLS} x_k\)., for
\(k = 1, ..., 4 x_{max} +1\). The regression equation based on the
principal axis was determined by using the \texttt{princomp} function in
the stats package in base R to obtain the rotation of the coordinate
axes from the first principal component (direction which captures the
most variance). The estimated slope, \(\hat\beta_{1,PCA}\), is
determined by the ratio of the axis rotation in y and axis rotation in x
of the first principal component with the y-intercept,
\(\hat\beta_{0,PCA}\) calculated by the point-slope equation of a line
using the mean of of the simulated points, \((\bar x_i, \bar y_i)\).
Fitted values, \(\hat y_{k,PCA}\), are then obtained every 0.25
increment across the domain from the PCA regression equation,
\(\hat y_{k,PCA} = \hat\beta_{0,PCA} + \hat\beta_{1,PCA} x_k\).

@@

{\centering \includegraphics[width=0.8\linewidth,]{Eye-Fitting-Straight-Lines-in-the-Modern-Era_files/figure-latex/ols-vs-pca-example-1}

@@

\caption{ Comparison between an OLS regression line which minimizes the vertical distance of points from the line and a regression line based on the principal axis which minimizes the Euclidean distance of points (orthogonal) from the line.}\label{fig:ols-vs-pca-example}
\end{figure}

\hypertarget{residual-trends}{%
\subsection{Residual Trends}\label{residual-trends}}

For each participant, the final data set used for analysis contains
\(x_{ijk}, y_{ijk,drawn}, \hat y_{ijk,OLS}\), and \(\hat y_{ijk,PCA}\)
for parameter choice \(i = 1,2,3,4\), j = \(1,...N_{participant}\), and
\(x_{ijk}\) value \(k = 1, ...,4 x_{max} + 1\). Using both a linear
mixed model and a generalized additive mixed model, comparisons of
vertical residuals in relation to the OLS fitted values
(\(e_{ijk,OLS} = y_{ijk,drawn} - \hat y_{ijk,OLS}\)) and PCA fitted
values (\(e_{ijk,PCA} = y_{ijk,drawn} - \hat y_{ijk,PCA}\)) were made
across the domain. \cref{fig:eyefitting-example-plot} displays an
example of all three fitted trend lines for parameter choice F. Data
used in the analyses are available to be downloaded from GitHub
\href{https://github.com/earobinson95/Eye-Fitting-Straight-Lines-in-the-Modern-Era/raw/main/data/youdrawit-eyefitting-model-data.csv}{here}.

@@

{\centering \includegraphics[width=0.5\linewidth,]{Eye-Fitting-Straight-Lines-in-the-Modern-Era_files/figure-latex/eyefitting-example-plot-1}

@@

\caption{Illustrates the data associated with and collected for one 'You Draw It' task plot. Trend-lines include the participant drawn line (dashed black), the OLS regression line (solid steelblue) and the PCA regression line based on the principal axis (solid orange).}\label{fig:eyefitting-example-plot}
\end{figure}

\hypertarget{linear-trend-constraint}{%
\subsubsection{Linear Trend Constraint}\label{linear-trend-constraint}}

Using the \texttt{lmer} function in the lme4 package \citep{lme4}, a
linear mixed model (LMM) is fit separately to the OLS residuals and PCA
residuals, constraining the fit to a linear trend. Parameter choice,
\(x\), and the interaction between \(x\) and parameter choice were
treated as fixed effects with a random participant effect accounting for
variation due to participant. The LMM equation for each fit (OLS and
PCA) is given by: \begin{equation}
y_{ijk,drawn} - \hat y_{ijk,fit} = e_{ijk,fit} = \left[\gamma_0 + \alpha_i\right] + \left[\gamma_{1} x_{ijk} + \gamma_{2i} x_{ijk}\right] + p_{j} + \epsilon_{ijk}
\end{equation} \noindent where

@@

Constraining the residual trend to a linear fit,
\cref{fig:eyefitting-lmer-residualplots} shows the estimated trend line
of the residuals between the participant drawn points and fitted values
for both the OLS regression line and PCA regression line. Estimated
residual trend lines are overlaid on the observed individual participant
residuals. Results indicate the estimated trends of PCA residuals
(orange) appear to align closer to the \(y=0\) horizontal (dashed) line
than the OLS residuals (blue). In particular, this trend is more
prominent in parameter choices with large variances (F and N). These
results are consistent to those found in \citet{mosteller1981eye}
indicating participants fit a trend-line closer to the estimated
regression line with the slope of based on the first principal axis than
the estimated OLS regression line thus, providing support for ``ensemble
perception''.

@@

\hypertarget{smoothing-spline-trend}{%
\subsubsection{Smoothing Spline Trend}\label{smoothing-spline-trend}}

Eliminating the linear trend constraint, the \texttt{bam} function in
the mgcv package \citep{mgcv1, mgcv2, mgcv3, mgcv4, mgcv5} is used to
fit a generalized additive mixed model (GAMM) separately to the OLS
residuals and PCA residuals to allow for estimation of smoothing
splines. Parameter choice was treated as a fixed effect with no
estimated intercept and a separate smoothing spline for \(x\) was
estimated for each parameter choice. A random participant effect
accounting for variation due to participant and a random spline for each
participant accounted for variation in spline for each participant. The
GAMM equation for each fit (OLS and PCA) residuals is given by:
\begin{equation}
y_{ijk, drawn} - \hat y_{ijk, fit} = e_{ijk,fit} = \alpha_i + s_{i}(x_{ijk}) + p_{j} + s_{j}(x_{ijk})
\end{equation} \noindent where

@@

Allowing for flexibility in the residual trend,
\cref{fig:eyefitting-gamm-residualplots} shows the estimated trend line
of the residuals between the participant drawn points and fitted values
for both the OLS regression line and PCA regression line. Estimated
residual trends are overlaid on the observed individual participant
residuals. The results of the GAMM align with those shown in
\cref{fig:eyefitting-lmer-residualplots} providing support that for
scatter-plots with more noise (F and N), estimated trends of PCA
residuals (orange) appear to align closer to the \(y=0\) horizontal
(dashed) line than the OLS residuals (blue). By fitting smoothing
splines, we can determine whether participants naturally fit a straight
trend-line to the set of points or whether they deviate throughout the
domain. In particular, in scatter-plots with smaller variance (S and V),
we can see that participants began at approximately the correct starting
point then deviated away from the fitted regression lines and correcting
for their fit toward the end of their trend-line. In scatter-plots with
larger variance (F and N), participants estimated their starting value
in the extreme direction of the OLS regression line based on the
increasing or decreasing trend but more accurately represented the
starting value of the PCA regression line. As participants continued
their trend-line, they crossed through the OLS regression line
indicating they estimated the slope in the extreme direction. These
results provide further insight into the curvature humans perceive in a
set of points.

@@

The intent of research was to adapt `You Draw It' from the New York
Times feature as a tool and method for testing graphics The study
conducted by \citet{mosteller1981eye} was replicated using the `You Draw
It' method providing support for the validity of the tool. Our results
indicated that when shown points following a linear trend, participants
tended to fit a line closer to a regression based on the principal axis
over the ordinary least squares regression line. This trend was most
prominent when shown data simulated with larger variances. These results
indicate that humans perform ``ensemble perception'' in a statistical
graphic setting as participants minimized the distance from the their
regression line over both the \(x\) and \(y\) axis simultaneously. In
addition, we allowed participants to draw trend lines that deviated from
a straight line providing insight into the curvature the human eye
perceives in a set of points.

@@

This study provided a basis for the use of `You Draw It' as a tool for
testing graphics as well as provided support for ``ensemble perception''
in statistical graphics. Further investigation is necessary to implement
the method in non-linear settings and with real data. This tool could
also be used to evaluate human ability to extrapolate data from trends.
In the future, an R package designed for easy implementation of `You
Draw It' task plots will help make this tool accessible to other
researchers.
